ODGI: scalable tools for pangenome graphs

### 2021-06-02 manuscript/main.tex
edits

    Thanks to advances in sequencing technology, new full-length
    genome assemblies are produced at a high rate, including efforts
    from the Human Pangenome Reference Consortium (HPRC) and
    telomere-to-telemere projects, e.g.~\citep{Miga:2020}. In
    principle, a pangenome can model a full set of genomic elements of
    individuals in a given species or clade. Pangenomics contrasts
    with reference genome based approaches which relate sequences to a
    single consensus model of the genome~\citep{Eizenga:2020}. A
    pangenome can elegantly be represented by a graph data structure
    incorporating sequences as nodes and their connections as
    edges. These nodes are shared for identical sequences, such as
    homologs, paralogs and orthologs. So-called paths can represent
    reads, contigs, haplotypes, individual chromosomes or part
    thereof, that run through the nodes. A bi-directed graph can even
    contain both strands of DNA as paths. A graph-based approach is
    particularly relevant for exploring complex variation and highly
    repetitive regions present which characterises the vertebrate
    genomes.

@@

@@

    \indent \topic{odgi build, view \& validate} pangenome graphs can be
    stored in the textual Graphical Fragment Assembly
    format~\citep{GFA}. \cmd{odgi build} and \cmd{odgi view}
    convert graphs from GFA to the \odgi\ format and \textit{vice versa}.
    \cmd{odgi validate} ascertains the graph data is correct, e.g.,
    it validates paths with respect to the graph topology.

    \topic{odgi viz, sort, draw \& layout} pangenome visualisation
    provides convenient insight into genomic variation. \cmd{odgi viz}
    generates a linearized representation of the pangenome (see Fig.~\ref{fig:1}a)
    and is capable of handling full length human chromosomes. \cmd{odgi sort}
    reorganises node order and simplifies the graph by applying several
    algorithms, including our novel 'path-guided stochastic gradient descent'
    algorithm. \cmd{odgi layout} generates a layout file with the X-Y
    coordinates for all nodes and \cmd{odgi draw} draws it, bringing
    the pangenome representation in two dimensions.

    \topic{odgi stats, bin, depth \& degree} pangenome numerical
    counts provide insight into complexity in other
    ways. \cmd{odgi stats} returns the number of nodes,
    edges, paths, and graph length. \cmd{odgi bin} summarises path
    information into bins of a specified size, thereby enabling a
    summarised view of large genomes. \cmd{odgi depth} and
    \cmd{odgi degree} return the node depth and degree as defined by
    user provided criteria. These methods allow the detection of complex
    genomic regions generated by highly repetitive sequences.

@@

@@

    % \vspace*{+12pt}
    % \textit{Financial Support}: none declared.

    \textit{Conflict of Interest}: none declared.
    \vspace*{+24pt}

        %TODO to repeat on a bigger machine to avoid vg crashing
    \begin{table}[h!]
        \caption{ Example of performance improvement between
            \cmd{odgi} and \cmd{vg}~\citep{vgtools} equivalent
            commands with output file sizes. The centromere region was
            extracted from CHR 6 of the HPRC year one assembly, i.e.,
            88~haploid, phased human genome assemblies from
            44~individuals plus the chm13 cell line and GRCh38
            reference genomes stored in 4.3GB~GFA format~\citep{GFA}.
            %
            First GFA was to converted to xg and next to \odgi\ and
            \vg\ formats respectively.  \odgi's path implementation
            can process paths in parallel, outperforming \vg. For the
            same reason \cmd{odgi extract} is $20\times$ faster than
            \cmd{vg chunk} extracting a portion from the full
            graph. \FIXME: vg never finished.
            %
            All timings were performed on a server class machine
            on SSD MODEL with X GB of RAM and 48 CPU cores (Y x 12 core
        AMD Opteron) Processor MODEL @ 3.3 GHz with 8MB L2
        Cache.\FIXME}
        \begin{tabular}{{@{}lllll@{}}}
          \toprule
          Operation & Command & Runtime & Memory  & File size     \\
                    &      & (mm:ss) & (GB) & (GB)           \\
          \midrule
          Covert GFA to native format                         &       &       &                   \\
          & \cmd{odgi build}      & 1:35  & 10.39 & 5.4               \\
          & \cmd{vg convert}      & 8:14  & 28.43 & 6.1               \\
          Extract subgraph                         &       &       &                   \\
          & \cmd{odgi extract}    & 1:15  & 9.43  & 2.7               \\
          & \cmd{vg chunk}        & 21:09 & 59.33 & 1.7               \\
          \botrule
        \end{tabular}
        \label{tab:02}

    \end{table}

@@

%    \vspace*{-12pt}

>>>>

    Thanks to advances in sequencing technology, new full-length
    genome assemblies are produced at a high rate, including efforts
    from the Human Pangenome Reference Consortium (HPRC) and
    telomere-to-telemere projects, e.g.~\citep{Miga:2020}. In
    principle, a pangenome can model a full set of genomic elements of
    individuals in a given species or clade. Pangenomics contrasts
    with reference genome based approaches which relate sequences to a
    single consensus model of the genome~\citep{Eizenga:2020}. A
    pangenome can elegantly be represented by a graph data structure
    incorporating sequences as nodes and their connections as
    edges. These nodes are shared for identical sequences, such as
    homologs, paralogs and orthologs. So-called paths can represent
    reads, contigs, haplotypes, individual chromosomes or part
    thereof, that run through the nodes. A bi-directed graph can even
    contain both strands of DNA as paths. A graph-based approach is
    particularly relevant for exploring complex variation and highly
    repetitive regions, such as the Major Histocompatibility Complex
    (MHC, Fig.~\ref{fig:1}).

@@

    \vspace{-0.15in}

@@

    \indent \topic{odgi build, view \& validate} pangenome graphs can be
    stored in the textual Graphical Fragment Assembly
    format~\citep{GFA}. \cmd{odgi build} and \cmd{odgi view}
    convert graphs from GFA to the \odgi\ format and \textit{vice versa}.
    \cmd{odgi validate} ascertains the graph data is correct.

    \topic{odgi viz, sort, draw \& layout} pangenome visualisation
    provides convenient insight into genomic variation. \cmd{odgi viz}
    generates a linearized representation of the pangenome (see Fig.~\ref{fig:1}a)
    and is capable of handling full length human chromosomes. \cmd{odgi sort}
    reorganises node order and simplifies the graph by applying several
    algorithms, including our novel 'path-guided stochastic gradient descent'
    algorithm. \cmd{odgi layout} generates a layout file with the X-Y
    coordinates for all nodes and \cmd{odgi draw} creates the figure.

    \topic{odgi stats, bin, depth \& degree} pangenome numerical
    counts provide insight into complexity in other
    ways. \cmd{odgi stats} returns the number of nodes,
    edges, paths, and graph length. \cmd{odgi bin} summarises path
    information into bins of a specified size, thereby enabling a
    summarised view of large genomes. \cmd{odgi depth} and
    \cmd{odgi degree} compute node depth and degree as defined by
    user provided criteria. These methods allow the detection of complex
    genomic regions generated by highly repetitive sequences.

@@

    \input{table}

@@

    \textit{Financial Support}: \FIXME\ Garrison \& Prins.
    % \vspace*{+12pt}

    \textit{Conflict of Interest}: none declared.

    % \vspace*{+12pt}

@@

%    \vspace*{-12pt}
    \vspace{-0.15in}
